---
title: "Machine Learning with Iris Data Set"
author: "Amyel Dale Cero"
date: "2025-09-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# R language equivalent of "Iris Flower" data analysis and machine learning

## 1. Install packages

```{r, eval = FALSE}

# Install packages

install.packages("RColorBrewer")  # Set a custom color palette like seaborn's 'husl'
install.packages("reshape2")     # For reshaping dataset

# MODELING LIBRARIES (similar to scikit-learn estimators)

install.packages("rsample")      # For separating training and testing sets
install.packages("caret")         # For ML: train/test split, training models, CV, accuracy, etc.

install.packages("nnet")          # For multinomial logistic regression
install.packages("e1071")         # For SVM and Naive Bayes
install.packages("kernlab")
install.packages("naivebayes")    # For Naive Bayes
install.packages("MASS")          # For LDA (Linear Discriminant Analysis)
install.packages("rpart")         # For Decision Trees
install.packages("rpart.plot")
install.packages("class")         # For KNN
install.packages("klaR")          # Alternative Naive Bayes
install.packages("nnet")          # Neural networks / multinomial logistic regression (optional)

```

## 2. Load the libraries

```{r}

# Load libraries

library(dplyr)         
library(tidyr)         
library(readr)         
library(ggplot2)
library(RColorBrewer)
library(reshape2)
library(rsample)
library(caret)
library(nnet)
library(MASS)
library(class)
library(rpart)
library(rpart.plot)
library(naivebayes)
library(e1071)
library(kernlab)

```

## 3. Get the data

```{r}

# Get the data from GitHub
# Note that Iris data set is also built-in in R

# Define the URL
url <- "https://raw.githubusercontent.com/jbrownlee/Datasets/master/iris.csv"

# Define column names
col_names <- c("sepal-length", "sepal-width", "petal-length", "petal-width", "class")

# Read the CSV from the URL with custom column names
dataset <- readr::read_csv(url, col_names = col_names)

# View the first few rows
head(dataset)

```

```{r}

# Check the structure
# Let us use the version of the data that was downloaded from the web

str(dataset)

```

```{r}

# Check the summary of the dataset
summary(dataset)

```

```{r}

# Count the number of occurrences

dataset %>%
  count(class)

```

## 4. Clean the data

```{r}

# Rename columns to avoid hyphens that may not be compatible with some ggplot functions

dataset_clean <- dataset %>%
  rename(
    sepal_length = `sepal-length`,
    sepal_width = `sepal-width`,
    petal_length = `petal-length`,
    petal_width = `petal-width`
  )

```

```{r}

# Ensure class is a factor
dataset_clean$class <- as.factor(dataset_clean$class)

```


## 5. Create  violin plots

```{r}

# Violin plot for Sepal Length

ggplot(dataset_clean, aes(x = class, y = sepal_length,
                          fill = class)) +
  geom_violin(trim = FALSE, fill = "skyblue") +
  stat_summary(fun = median, geom = "point", color = "red", size = 2) +
  ggtitle("Sepal Length by Class")

```


```{r}

# Violin plot for Sepal Width

ggplot(dataset_clean, aes(x = class, y = sepal_width,
                          fill = class)) +
  geom_violin(trim = FALSE, fill = "lightgreen") +
  stat_summary(fun = median, geom = "point", color = "red", size = 2) +
  ggtitle("Sepal Width by Class")

```

```{r}

# Violin plot for Petal Length

ggplot(dataset_clean, aes(x = class, 
                          y = petal_length,
                          fill = class)) +
  geom_violin(trim = FALSE, fill = "salmon") +
  stat_summary(fun = median, geom = "point", color = "red", size = 2) +
  ggtitle("Petal Length by Class")

```

```{r}

# Violin plot for Petal Width

ggplot(dataset_clean, aes(x = class, 
                          y = petal_width,
                          fill = class)) +
  geom_violin(trim = FALSE, fill = "plum") +
  stat_summary(fun = median, geom = "point", color = "red", size = 2) +
  ggtitle("Petal Width by Class")

```

## 6. Create a correlation heatmap

```{r}

# Create a correlation heatmap

# Compute correlation matrix
cor_matrix <- cor(dataset_clean[, 1:4])

# Melt for ggplot
cor_melt <- melt(cor_matrix)

# Plot correlation heatmap
ggplot(cor_melt, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile() +
  geom_text(aes(label = round(value, 2)), color = "white") +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  labs(title = "Correlation Heatmap of Iris Features", x = "", y = "") +
  theme_minimal()

```

## 7. Prepare data for modeling

```{r}

# Drop 'class' column to get predictors (X)
x <- dataset_clean[, setdiff(names(dataset_clean), "class")]

# Extract 'class' column as target (y)
y <- dataset_clean[["class"]]

# Print shapes (dimensions)
cat(sprintf("X shape: %d x %d | y shape: %d\n", nrow(x), ncol(x), length(y)))
```

## 8. Set training and testing sets

```{r}

# Combine x and y into one dataset if needed
data_all <- cbind(x, class = y)

# Create split object
set.seed(1)
split_obj <- initial_split(data_all, prop = 0.6, strata = class)

# Training and testing sets
train_data <- training(split_obj)
test_data  <- testing(split_obj)

# Optionally extract x/y again
x_train <- train_data[, setdiff(names(train_data), "class")]
y_train <- train_data$class

x_test  <- test_data[, setdiff(names(test_data), "class")]
y_test  <- test_data$class

```

## 9. Compare models

```{r}

# Define training control with 10-fold stratified CV
train_control <- trainControl(
  method = "cv",      # cross-validation
  number = 10,        # 10 folds
  classProbs = FALSE, # no probabilities needed
  summaryFunction = defaultSummary  # default metrics like Accuracy
)

# Model list with caret method names
model_list <- list(
  LR   = "multinom",     # Multinomial Logistic Regression
  LDA  = "lda",          # Linear Discriminant Analysis
  KNN  = "knn",          # K-Nearest Neighbors
  CART = "rpart",        # Decision Tree
  NB   = "naive_bayes",  # Naive Bayes (needs 'naivebayes' package)
  SVC  = "svmLinear"     # Support Vector Classifier
)

```

```{r}

# Store results
results <- list()

# Loop through and train each model
for (model_name in names(model_list)) {
  cat("\nTraining:", model_name, "\n")
  
  fit <- train(
    x = x_train,
    y = y_train,
    method = model_list[[model_name]],
    trControl = train_control,
    metric = "Accuracy"
  )
  
  results[[model_name]] <- fit
  
  # Find the best tuning parameters
  best_params <- fit$bestTune
  
  # Find the row index in results that matches bestTune (works even if bestTune is empty for no tuning)
  if (nrow(best_params) == 0) {
    # No tuning parameters, just take the first row
    idx <- 1
  } else {
    # Match bestTune values in results
    idx <- which(apply(fit$results[, names(best_params), drop=FALSE], 1, function(row) all(row == best_params)))
  }
  
  # Extract accuracy and std dev for best params
  acc <- fit$results$Accuracy[idx]
  std <- fit$results$AccuracySD[idx]
  
  cat(sprintf("%s: Accuracy = %.4f (SD = %.4f)\n", model_name, acc, std))
}

```

## 10. Try for a specific model

```{r}

# Train SVM model (linear kernel)
model <- train(
  x = x_train,
  y = y_train,
  method = "svmLinear",
  trControl = trainControl(method = "none")  # no resampling, just train
)

# Predict on test data
prediction <- predict(model, newdata = x_test)

# View predictions
print(prediction)

```

```{r}

# Calculate accuracy
accuracy <- sum(prediction == y_test) / length(y_test)
cat(sprintf("Test Accuracy: %.4f\n", accuracy))

# Classification report (confusion matrix and stats)
conf_mat <- confusionMatrix(prediction, y_test)
print(conf_mat)

```

## 11. Linear regression workflow

```{r}

# 1. Load the Iris dataset
data(iris)

```

```{r}

# 2. Select features
X <- iris[["Petal.Length"]]  # Independent variable
y <- iris[["Petal.Width"]]   # Dependent variable

```

```{r}

# 3. Split into training and testing sets
set.seed(42)
train_index <- sample(seq_len(nrow(iris)), size = 0.8 * nrow(iris))
X_train <- X[train_index]
X_test  <- X[-train_index]
y_train <- y[train_index]
y_test  <- y[-train_index]

```


```{r}

# 4. Train linear regression model
model <- lm(y_train ~ X_train)

```

```{r}

# 5. Make predictions
y_pred <- predict(model, newdata = data.frame(X_train = X_test))

```

```{r}

# 6. Print model details
cat("Linear Regression Model:\n")
cat(sprintf("Coefficient (m): %.4f\n", coef(model)[2]))
cat(sprintf("Intercept (c): %.4f\n", coef(model)[1]))

```

```{r}

# 7. Calculate R-squared
r_squared <- summary(model)$r.squared
cat(sprintf("R-squared (R²): %.4f\n", r_squared))

```

```{r}

# 8. Visualization

plot_data <- data.frame(
  Petal.Length = X_test,
  Petal.Width = y_test,
  Predicted = y_pred
)

ggplot(plot_data, aes(x = Petal.Length, y = Petal.Width)) +
  geom_point(color = "blue", size = 2, alpha = 0.6) +
  geom_line(aes(y = Predicted), color = "red", size = 1.2) +
  labs(
    title = "Linear Regression on Iris Dataset",
    x = "Petal Length (cm)",
    y = "Petal Width (cm)"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_y_continuous(sec.axis = dup_axis(name = NULL)) +
  guides(color = "none")

```

## 12. Support Vector Classifier (SVC) with a linear kernel

```{r}

# 1. Load the Iris dataset
data(iris)

```

```{r}

# 2. Keep only the first two features for visualization
iris_subset <- iris[, c("Sepal.Length", "Sepal.Width", "Species")]

```

```{r}

# 3. Split into train and test sets (80/20)
set.seed(42)
train_index <- createDataPartition(iris_subset$Species, p = 0.8, list = FALSE)
train_data <- iris_subset[train_index, ]
test_data  <- iris_subset[-train_index, ]

```

```{r}

# 4. Train the SVM model with a linear kernel
model <- svm(Species ~ Sepal.Length + Sepal.Width, 
             data = train_data, 
             kernel = "linear", 
             cost = 1.0, 
             scale = TRUE)
```

```{r}

# 5. Make predictions and evaluate
pred <- predict(model, newdata = test_data)

cat("Support Vector Classifier Model Performance:\n")
accuracy <- mean(pred == test_data$Species)
cat(sprintf("Accuracy: %.2f\n\n", accuracy))

```

```{r}

# Classification report
conf_mat <- confusionMatrix(pred, test_data$Species)
print(conf_mat)

```

```{r}

# 6. Visualize the decision boundaries
# Create a grid to evaluate model
x_min <- min(iris_subset$Sepal.Length) - 0.5
x_max <- max(iris_subset$Sepal.Length) + 0.5
y_min <- min(iris_subset$Sepal.Width) - 0.5
y_max <- max(iris_subset$Sepal.Width) + 0.5

```

```{r}

# Resolution of the grid
h <- 0.02
grid <- expand.grid(
  Sepal.Length = seq(x_min, x_max, by = h),
  Sepal.Width  = seq(y_min, y_max, by = h)
)

```

```{r}

# Predict the class for each point in the grid
grid$Species <- predict(model, newdata = grid)

```

```{r}

# 7. Plot
ggplot() +
  geom_tile(data = grid, aes(x = Sepal.Length, y = Sepal.Width, fill = Species), alpha = 0.3) +
  geom_point(data = iris_subset, 
             aes(x = Sepal.Length, y = Sepal.Width, color = Species),
             size = 2, shape = 21, stroke = 1) +
  scale_fill_manual(values = c("#D55E00", "#0072B2", "#009E73")) +
  scale_color_manual(values = c("#D55E00", "#0072B2", "#009E73")) +
  labs(title = "SVC Decision Boundary on Iris Dataset",
       x = "Sepal Length (cm)",
       y = "Sepal Width (cm)") +
  theme_minimal() +
  theme(legend.position = "right")

```

## 13. Linear Discriminant Analysis (LDA)

```{r}

# 1. Load the Iris dataset
data(iris)

```

```{r}

# 2. Set features (X) and target (y)
X <- iris[, 1:4]            # All 4 numeric features
y <- iris$Species           # Target variable

```

```{r}

# 3. Split data into training and testing sets (70/30)
set.seed(42)
train_index <- createDataPartition(y, p = 0.7, list = FALSE)
X_train <- X[train_index, ]
y_train <- y[train_index]
X_test  <- X[-train_index, ]
y_test  <- y[-train_index]

```

```{r}

# 4. Train LDA model (and transform training data)
lda_model <- lda(Species ~ ., data = iris[train_index, ])
X_train_lda <- predict(lda_model, X_train)$x

```

```{r}

# 5. Predict on test set
lda_pred <- predict(lda_model, X_test)
y_pred <- lda_pred$class

```

```{r}

# 6. Evaluate performance
cat("Linear Discriminant Analysis (LDA) Model Performance:\n")
accuracy <- mean(y_pred == y_test)
cat(sprintf("Accuracy: %.2f\n\n", accuracy))

conf_matrix <- confusionMatrix(y_pred, y_test)
print(conf_matrix)

```

```{r}

# 7. Visualize the LDA projection (entire dataset)
lda_all <- lda(Species ~ ., data = iris)
lda_values <- predict(lda_all)$x  # Projected values

# Combine with original labels for plotting
plot_data <- data.frame(lda_values, Species = iris$Species)

```

```{r}

# 8. Plot
ggplot(plot_data, aes(x = LD1, y = LD2, color = Species)) +
  geom_point(size = 3, alpha = 0.8) +
  labs(
    title = "LDA of Iris Dataset",
    x = "Linear Discriminant 1",
    y = "Linear Discriminant 2"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_color_viridis_d(option = "viridis")

```
## 14. K-Nearest Neighbors (KNN)

```{r}

# 1. Load the Iris dataset
data(iris)

```

```{r}

# Define features and target
X <- iris[, 1:4]            # All four features
y <- iris$Species           # Target variable
feature_names <- names(X)

```

```{r}

# 3. Split the dataset into training and testing sets (70/30)
set.seed(42)
train_index <- createDataPartition(y, p = 0.7, list = FALSE)
X_train <- X[train_index, ]
X_test  <- X[-train_index, ]
y_train <- y[train_index]
y_test  <- y[-train_index]

```

```{r}

# 4. Train and apply KNN model (k = 5)
k <- 5
y_pred <- knn(train = X_train, test = X_test, cl = y_train, k = k)

```

```{r}

# 5. Evaluate model performance
cat("K-Nearest Neighbors (KNN) Model Performance:\n")
accuracy <- mean(y_pred == y_test)
cat(sprintf("Accuracy: %.2f\n\n", accuracy))

# Confusion matrix and classification metrics
conf_matrix <- confusionMatrix(y_pred, y_test)
print(conf_matrix)

```

```{r}

# 6. Visualization using first two features
# We'll use original full dataset and plot Sepal.Length vs. Sepal.Width

ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point(size = 3, alpha = 0.8, shape = 21, stroke = 1) +
  labs(
    title = "Iris Dataset: Sepal Length vs. Sepal Width",
    x = "Sepal Length (cm)",
    y = "Sepal Width (cm)",
    color = "Species"
  ) +
  theme_minimal() +
  scale_color_viridis_d(option = "viridis") +
  theme(plot.title = element_text(hjust = 0.5))

```

## 15. CART (Decision Tree)

```{r}

# 1. Load the Iris dataset
data(iris)

```

```{r}

# 2. Set seed for reproducibility
set.seed(42)

```

```{r}

# 3. Split the data into training and testing sets (70/30)
train_index <- createDataPartition(iris$Species, p = 0.7, list = FALSE)
train_data <- iris[train_index, ]
test_data  <- iris[-train_index, ]

```

```{r}

# 4. Train a CART model (Decision Tree with max depth = 3)
cart_model <- rpart(Species ~ ., data = train_data, method = "class",
                    control = rpart.control(maxdepth = 3))

```

```{r}

# 5. Make predictions
y_pred <- predict(cart_model, test_data, type = "class")

```

```{r}

# 6. Evaluate the model
cat("Classification and Regression Tree (CART) Model Performance:\n")
accuracy <- mean(y_pred == test_data$Species)
cat(sprintf("Accuracy: %.2f\n\n", accuracy))

conf_matrix <- confusionMatrix(y_pred, test_data$Species)
print(conf_matrix)

```

```{r}

# 7. Visualize the decision tree
rpart.plot(cart_model,
           type = 2,             # Boxed nodes
           extra = 104,          # Show probabilities + percentages
           under = TRUE,
           faclen = 0,           # Show full factor names
           main = "Decision Tree for Iris Dataset")

```

## 16. Naive Bayes classifier

```{r}

# 1. Load the iris dataset
data(iris)

```

```{r}

# 2. Optional: get feature and class names (like Python's `feature_names` and `target_names`)
feature_names <- names(iris)[1:4]
target_names <- levels(iris$Species)

```

```{r}

# 3. Split the dataset into training and testing sets (70/30)
set.seed(42)
train_index <- createDataPartition(iris$Species, p = 0.7, list = FALSE)
train_data <- iris[train_index, ]
test_data  <- iris[-train_index, ]

```

```{r}

# 4. Train a Gaussian Naive Bayes model
nb_model <- naiveBayes(Species ~ ., data = train_data)

```

```{r}

# 5. Make predictions
y_pred <- predict(nb_model, newdata = test_data)

```

```{r}

# 6. Evaluate the model
cat("Naive Bayes (NB) Model Performance:\n")
accuracy <- mean(y_pred == test_data$Species)
cat(sprintf("Accuracy: %.2f\n\n", accuracy))

conf_matrix <- confusionMatrix(y_pred, test_data$Species)
print(conf_matrix)

```

```{r}

# 7. Visualize the data using the first two features (Sepal.Length vs Sepal.Width)
ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_viridis_d() +
  labs(
    title = "Iris Dataset: Sepal Length vs. Sepal Width",
    x = feature_names[1],
    y = feature_names[2],
    color = "Species"
  ) +
  theme_minimal() +
  theme(
    text = element_text(size = 14),
    legend.title = element_text(face = "bold")
  )

```

