---
title: "Basic R Programming (Additional Concepts)"
author: "Amyel Dale Cero"
date: "2025-10-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. If-else statements

## 1.1. Logic operators

### 1.1.1. Examples

```{r}

# Greater than
2005 > 2001


```

```{r}

# Equal
2005 = 2005

```

This will produce an error because the correct logical operator for is ==

```{r}

# Correct equal sign
2005 == 2005

```

```{r}

2005 != 2008

# Why is this TRUE?

```

### 1.1.2. Examples of OR (\|) and AND (&)

Analyze why the following codes produce TRUE or FALSE:

```{r}

(2005 > 2008) | (2003 < 2010)

```

```{r}

(2005 > 2008) & (2003 < 2010)

```

```{r}

!(2005 > 2008) & (2003 < 2010)

```

```{r}

!(2005 > 2008) & (2003 < 2010) & (1801 > 1819)

```

## 1.2. Basic if statements

```{r}

Test_1 <- 90

if(Test_1 > 90){
  Grade <- "A"
  }

print(Grade)
```

What if we change the value of Test_1 to 90?

## 1.3. Nested statements

```{r}

# Initialize variable A
A <- 10

```

```{r}

# Create a nested if-else statement

if(A > 20){
  B <- 3
} else if (A > 10){
  B <- 6
} else {
  B <- 9
}

# Note that there is a very specific syntax for this.
# The closing brace of conditional evaluation must be on the same line as else
  
```

```{r}

print(B)

```

Using the example above from the basic if statement:

```{r}

Test1 <- 93

if(Test1 > 90){
  Grade <- 'A'
} else if ( Test1 > 80) {
  Grade <- 'B'
} else if (Test1 > 70) {
  Grade <- 'C'
} else {
  Grade <- 'D'
}

```

```{r}

print(Grade)

```

Using the example on what grade school students must wear depending on air temperature:

```{r}

T <- 50 # must initialize T variable

if (T < 10) {
  print("No recess")
}else if( T< 25 ) {
  print("Bring snowpants, hat, and gloves")
}else if( T< 40) {
  print("Bring hat and gloves")
}else {
  print("Nothing special needed.")}

```

Test out the code by changing the T.

## 1.4. Switch function

Example 1: Match variables and then return the associated variable

```{r}

# Switching color to a corresponding fruit

switch("green",      # first argument = input to be switched
       "green" = "cucumber", # second and onwards after comma = conditions
       "red" = "cherry tomato",
       "yellow" = "summer squash"
         )
```

Example 2: Indicate position and then retrieve the value from that position.

```{r}

switch(4,
       7, 14, 21, 28)

```

# 2. Loops

## 2.1. For loops

Example 1: Write a for loop that prints numbers 0 to 100 by 5

```{r}

for (i in 0:20){
  print (i * 5)
}

```

Example 2: Write a for loop that adds and prints numbers 0 to 100 by 5

```{r}

# Initialize variable that will hold current value at each step
Sum <- 0

for (i in 0:20){
  
  Sum <- Sum + (i * 5)
  print(Sum)
  
}

```

Example 3: Using character variables

```{r}

# Initialize a vector containing different breeds of dogs
dog_vector <- c("pug", "retriever", "terrier", "mastiff")

# Add a statement before each breed of the dog
for (dogs in dog_vector){
  print(paste("The breed of the month is", dogs))
}

```

## 2.2. While loops

Example 1: Construct a WHILE loop that stops when the counter reaches 50. Start at a value of 1. Print out each iteration.

```{r}

# Initialize variables
i <- 1 # counter

while (i < 51){
  
  i = i + 1
  print(i-1)
  
}

```

## 2.3. Using break

Example 1: Make a loop that goes through a series of random numbera and stops when the number is \> 150

```{r}

Length <- 1000

# Generate a vector of random numbers
A <- round(rnorm(Length) * 100)

# Make the loop
for (i in 1:Length){
  
  print(i)   # Print the number of the current iteration
  
  if (A[i] > 150) {
    print("I'm done")
    break
  }
}

```

# 3. Functions

Example 2: You find yourself frequently needing to convert between Fahrenheit and Celsius degree. Write a function that allows you to input Fahrenheit degrees and get out Celsius degrees.

Keep in mind that Celsius degrees are 32 degrees less and 5/9 of Fahrenheit degrees.

```{r}

# Name of the function = FarenToCel
# Input: temperature in F

FarenToCel <- function(Temp_in_F){
  
  Temp_in_C <- (Temp_in_F - 32) * 5/9    # What should be done to the input
  
}

# Initialize/source the function by running this chunk of code

```

Notice what changes in the Environment pane.

```{r}

# Use the function

FarenToCel(100)

```

Where is the result? Since there is no print() inside the function, the result has to be assigned to a variable outside the function.

```{r}

Temp_C <- FarenToCel(100)
Temp_C

```

Example 3: Write a function that calculates someone’s age in year 2025 when given their birth year. Then, write a line of code that calls the function.

```{r}

AgeCalc <- function(Input_Yr){
  
  Age <- 2025 - Input_Yr
  print(Age)
  
}
```

```{r}

AgeCalc(1990)

```

Notice that this one readily prints the answer because we included a print function inside the function.
