---
title: "Exercise"
author: "Amyel Dale Cero"
date: "2025-10-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Basic R programming concepts

**Instructions:**

Write a code that will solve the following. Add comments (\#) to explain your code, if needed.

# 1. Vectors and matrices in R

1.  Create a numeric vector of 5 numbers and assign it an R object named `num`

```{r}


```

2.  Create a character vector with names of 5 of your favorite fruits. Assign it to an R object named `fruit`

```{r}


```

3.  Access the 3rd element of `num`

```{r}


```

4.  Access the first three `fruit`

```{r}


```

# 2. If-else

Create a temperature variable called `temp`.

Write an `if-else` statement such that:

-   if the temperature is greater than 30, the code prints "It's hot!"

-   if the temperature is greater than or equal to 20, and less than or equal to 30, the code prints "It's warm."

-   otherwise, the code prints "It's cold."

```{r}


```

# 3. Loops

Write a for loop that prints number from 1 to 5

```{r}


```

Write a while loop that keeps adding 1 to a variable x until it reaches 10

```{r}


```

# 4. Functions

Create a function that computes for the number of days given the number of months. Assume that 1 month = 30 days.

```{r}


```

Show that the function works properly.

```{r}


```

# 5. Data frames in R

1.  Create a data frame named `inventory` with the following columns and their elements (5 each):

-   fund source = Region I, Region II, Region XI, DA-OSEC, PHilMech
-   rated power = 70, 42, 50, 55, 70
-   supplier = YANMAR, ADAMCO, FORD TRACT, AGRIMACH, INA

```{r}


```

2.  Convert the column of fund source from character to categorical.

```{r}

```

3.  Check if the column was converted properly.

```{r}

```

4.  Get the mean of the rated power

```{r}

```

5.  Add a new column: year funded = 2022, 2013, 2022, 2010, 2021

```{r}

```

6.  Check if the new column was added successfully

```{r}

```

7.  Check the distribution of rated power using a histogram. Make sure to add an appropriate title and x-axis label.

```{r}

```

# 6. File Input/Output and EDA concepts

Download the following .csv file:
https://github.com/ADLCero/Intro_to_R_training/blob/main/00_Data/Syr_MetData.csv

Save it where it should be saved inside your R Project.

Read in the .csv file into R and assign it to an R object called `weather`

```{r}


```

Print the first few rows of the data

```{r}

```

Determine how many rows and columns there are in the data

```{r}

```

Check the structure of the data

```{r}


```

Check some basic statistics about the data

```{r}


```

What can you notice?

Create a scatterplot matrix of the data set using only the numeric columns

```{r}


```

Create a heatmap using only the numeric columns of the data set.

Hint: There should be no NAs in the dataset for this to work properly. 
To remove NAs, we can use the function `na.omit()`.

Since this was not discussed yesterday, how can we know how `na.omit()` works without asking the resource person? :)

```{r}

# Get the numeric columns only


# Remove NAs


# Check if the NAs have been removed by comparing the number of rows w/ and w/o using na.omit


# Compute for the correlation matrix of all pairs of numeric variables


# Create the heat map

```

Write out the version of the data without NAs as "weather_na.csv" and save it the Output folder of your R Project.

```{r}


```

Check if it was indeed saved in the correct folder.

Congratulations, you are done! 

