---
title: "Exercise"
author: "Amyel Dale Cero"
date: "2025-10-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Basic R programming concepts

**Instructions:**

Write a code that will solve the following. Add comments (\#) to explain your code, if needed.

# 1. Vectors and matrices in R

1.  Create a numeric vector of 5 numbers and assign it an R object named `num`

```{r}

num <- c(7, 11, 17, 27, 1)
print(num)

```

2.  Create a character vector with names of 5 of your favorite fruits. Assign it to an R object named `fruit`

```{r}

fruit <- c("mango", "banana", "star apple", "persimmon", "cantaloupe")
print(fruit)

```

3.  Access the 3rd element of `num`

```{r}

num[3]

```

4.  Access the first three `fruit`

```{r}

fruit[1:3]

```

# 2. Data frames in R

1.  Create a data frame named `inventory` with the following columns and their elements (5 each):

<!-- -->

a.  fund source = Region I, Region II, Region XI, DA-OSEC, PHilMech

b.  rated power = 70, 42, 50, 55, 70

c.  supplier = YANMAR, ADAMCO, FORD TRACT, AGRIMACH, INA

```{r}

inventory <- data.frame(
  fund_source = c("Region I", "Region II", "Region XI", "DA-OSEC", "PHilMech"),
  rated_power = c(70, 42, 50, 55, 70),
  supplier = c("YANMAR", "ADAMCO", "FORD TRACT", "AGRIMACH", "INA")
  )

print(inventory)

```

2.  Convert the column on fund source from character to categorical.

```{r}

inventory$fund_source <- as.factor(inventory$fund_source)

```

3.  Check if the column was converted properly.

```{r}

# Approach 1:
str(inventory)

# Approach 2:
class(inventory$fund_source)

```

4.  Get the mean of the rated power

```{r}

mean(inventory$rated_power)

```

5.  Add a new column: year funded = 2022, 2013, 2022, 2010, 2021

```{r}

inventory$year_funded <- c(2022, 2013, 2022, 2010, 2021)

```

6.  Check if the new column was added successfully

```{r}

names(inventory)
head(inventory)
View(inventory)

```

7.  Check the distribution of rated power using a histogram. Make sure to add an appropriate title and x-axis label.

```{r}

hist(inventory$rated_power,
     main = "Rated Power of Inventoried Machineries",
     xlab = "rated power", 
     col = "orange", 
     border = "steelblue", 
     lwd = 4)

```

# 3. If-else

Create a temperature variable called `temp`.

Write an `if-else` statement such that:
- if the temperature is greater than 30, the code prints "It's hot!"
- if the temperature is greater than or equal to 20, and less than or equal to 30, the code prints "It's warm."
- otherwise, the code prints "It's cold."

```{r}

temp <- 20

if(temp > 30){
  print("It's hot!")
} else if (temp >= 20 & temp <= 30){
  print("It's warm.")
} else {
  print("It's cold.")
}

```

# 4. Loops

Write a for loop that prints number from 1 to 5

```{r}

for (i in 1:5){
  
  print(i)
  
}
```

Write a while loop that keeps adding 1 to a variable x until it reaches 10

```{r}

x <- 0
i <- 1

while (i < 11){
  
  x <- x + 1
  print(x)
  
  i <- i + 1
  
}
```

# 5. Functions

Create a function that computes for the number of days given the number of months. Assume that 1 month = 30 days.

```{r}

Number_days <- function(number_months){
  
  Total_days <- number_months * 30
  
  return(Total_days)
  
}
```

Show that the function works properly.

```{r}

Number_days(10)

```

# 6. File Input/Output

Download the following 